---
title: Powerlifting - Tidy Tuesday
author: Jay Achar
date: '2019-10-19'
slug: tt-19-41-powerlifting
categories:
  - dataviz
tags:
  - tidytuesday
  - R
description: 'Practicing with tidybayes'
featured: "gym-sign.jpeg"
featuredalt: "Gym-sign"
---
```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(naniar)  # visulising missing data
library(hrbrthemes)
library(tidybayes)   # dev version from github
library(skimr)

color7_palette <- c("#f20000", "#f94827", "#ff6b46", "#ff8966", "#ffa486", "#ffbea8", "#ffd8cb")
```
```{r import-data, include=FALSE}
# name file 
file <- "~/Dropbox/Data/tidy_tuesday/tt-19-41-powerlifting.rds"

# only download file if not previously saved
if (file.exists(file)) {
  ipf_lifts <- readRDS(file)
} else {
  ipf_lifts <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-08/ipf_lifts.csv")
  saveRDS(ipf_lifts, file = file)
}

```

This week's [#TidyTuesday](https://thomasmock.netlify.com/post/tidytuesday-a-weekly-social-data-project-in-r/){target="_blank"} topic is international powerlifting. While it's certainly not a topic I know much about or am particularly interested in, the description of the data suggests there's plenty to dig into. 

This is a short blog post describing my exploratory analysis with the intention of generating an attractive but also an informative final plot for my [Twitter feed](https://www.twitter.com/dr_jay_a){target="_blank"}. I've recently found the `tidybayes` package and wanted to try out some of the ggplot-compatible functions that it contains. 

## Data

Let's start by importing the data from the [Tidy Tuesday Github repo](https://github.com/rfordatascience/tidytuesday){target="_blank"} and taking a quick look at it's structure. 

```{r import-data-show, eval=FALSE}
ipf_lifts <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-08/ipf_lifts.csv")
```
```{r glimpse-data}
glimpse(ipf_lifts)
```
So, it looks like each lifter's efforts in different events are represented by a single row resulting in `r nrow(ipf_lifts)` rows...impressive. 

What about some summary statistics to get another view of this huge data set: 
```{r data-summary}
ipf_lifts %>% skim()

```

## Missing data

Is there much missing data in the most important variables? Let's choose some key variables and plot a simple summary of missingness. 

```{r missing-data, echo=FALSE}
p_missing_data <- ipf_lifts %>% 
  select(sex, age, bodyweight_kg, starts_with("best3"), date) %>% 
  naniar::miss_var_summary()
  
p_missing_data 
```

# Distribution

I"m interested to look more carefully at the maximum weights lifted within the data set, so I'll start by plotting their distributions:

```{r best-distribution, fig.height=12, fig.width=6, echo=FALSE}
p_distribution <- ipf_lifts %>% 
  select(starts_with("best3")) %>% 
  rename(Squat = "best3squat_kg", 
         Bench = "best3bench_kg",
         Deadlift = "best3deadlift_kg") %>% 
  pivot_longer(cols = everything(), names_to = "type", values_to = "weight") %>% 
  filter(!is.na(weight) & weight > 0) %>% 
  ggplot(aes(x = weight)) +
    geom_histogram(binwidth = 10) +
    facet_wrap(facets = ~ type, nrow = 3) +
  labs(title = "Distribution of maximum weights lifted",
       subtitle = "",
       x = "Maximum weight lifted (kgs)",
       y = "Count") +
  theme_ipsum() +
    theme(axis.ticks = element_blank())

p_distribution
```

With this number of data points, it's not surprising that the distribution is reasonably close to gaussian. 

I wonder who has the most records in the dataset:
```{r most-records}
ipf_lifts %>% 
  count(name) %>% 
  arrange(desc(n))
```
That's a lot of events!

Sex, age, age_class, bodyweight_kg and date seem self-explanatory. Looking at the codebook on the [Tidy Tuesday data page](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-10-08){target="_blank"} fills in the rest of the gaps.

## Temporal trends

Let's take a look at how many rows there are per unit time: 
```{r rows-by-year, echo=FALSE}

p_rows_time <- ipf_lifts %>% 
  count(date) %>% 
  ggplot(aes(x = date, y = n)) +
   geom_point(alpha = 0.4) +
   stat_smooth(method = "loess", se = FALSE, color = "red") +
  theme_ipsum() +
  labs(title = "Data rows by date",
       subtitle = "More data points for more recent events",
       x = "Date",
       y = "Number of data points") +
  theme(axis.title.x = element_text(size = 10,
                                    vjust = -2,
                                    hjust = 0.95),
        axis.title.y = element_text(size = 10, 
                                    vjust = 2,
                                    hjust = 0.95),
        axis.text.y = element_text(size = 9, color = "grey65"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
p_rows_time
```

That either suggests there are more events or that there are more participants at each event in recent years. The distribution does seem to be affected by some higher outliers in the last decade rather than there being a consistent upward trend. 

Are there other temporal relationships to consider? 

```{r gender-by-year, echo=FALSE}
p_gender_time <- ipf_lifts %>% 
  count(date, sex) %>% 
  ggplot(aes(x = date, y = n, group = sex)) +
    geom_point(aes(color = sex), alpha = 0.5) +
    stat_smooth(aes(color = sex), method = "loess", se = FALSE) +
theme_ipsum() +
  labs(title = "Data rows by date and gender",
       subtitle = "More men than women throughout the time period",
       x = "Date",
       y = "Number of data points", 
       color = "Gender") +
  theme(axis.title.x = element_text(size = 10,
                                    vjust = -2,
                                    hjust = 0.95),
        axis.title.y = element_text(size = 10, 
                                    vjust = 2,
                                    hjust = 0.95),
        axis.text.y = element_text(size = 9, color = "grey65"),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank())


p_gender_time
```

The temporal trend in the quantity of data seems reasonably consistent across gender where data quantity is consistently lower for women than men. 

How about bodyweight? Are lifters getting heavier or lighter over time? I'll plot the distributions of male powerlifters to check whether we can see any temporal change in medians or distributions. I'm using the [tidybayes package](https://mjskay.github.io/tidybayes/index.html){target="_blank"} for this which includes a number of geoms_* and stats_* within the Github development version. 

```{r bodyweight-decade, echo=FALSE}
p_bodyweight_decade <- ipf_lifts %>%
  # create decade variable
  mutate(decade = year(date) - (year(date) %% 10)) %>%
  filter(!is.na(bodyweight_kg)) %>% 
  filter(sex == "M") %>% 
  ggplot(aes(x = bodyweight_kg, y = decade)) +
   stat_intervalh(.width = c(.5, .75, .90, .99),
                   height = 5, show.legend = F) +
   scale_color_manual(values = rev(color7_palette[c(1, 3, 5, 7)])) +
  # rcartocolor::scale_color_carto_d(palette = "Peach") + 
   stat_halfeyeh(aes(as.numeric(bodyweight_kg), decade),
                  .width = 0, fill = color7_palette[7], alpha = 0.5,
                  size = 0.7, point_alpha = 1, height = 2.5) +
  scale_x_continuous(limits = c(0, 250)) +
  coord_flip() +
  labs(title = "Bodyweight distribution in men by decade",
       subtitle = "Wider distribution in recent decades suggests heavier outliers",
       y = "", 
       x = "Bodyweight (kgs)") +
  theme_ipsum() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.major.y = element_line(size = 0.1),
          axis.text.y = element_text(size = 9, color = "grey65"))
p_bodyweight_decade
```

Is there any trend in the maximum deadlift weights by time and gender? 

```{r deadlift-gender-time, echo=FALSE}
ipf_lifts %>% 
  mutate(decade = year(date) - (year(date) %% 10)) %>%
  mutate(gender = ifelse(sex == "M", "Male", 
                         ifelse(sex == "F", "Female", NA_character_))) %>% 
  filter(!is.na(decade) & !is.na(best3deadlift_kg)) %>% 
  filter(best3deadlift_kg > 0) %>% 
  ggplot(aes(x = decade, y = best3deadlift_kg)) +
    stat_interval(.width = c(0.5, 0.75, 0.90, 0.99),
                  show.legend = TRUE) +
    stat_dots(.width = 0, alpha = 0.7, height = 0.2, size = 0.2,
              position = position_nudge(x = 0.5),
              color = color7_palette[7]) +
  scale_color_manual(values = rev(color7_palette[c(1, 3, 5, 7)])) +
  facet_wrap(~ gender, nrow = 1) +
  labs(title = "Distribution of maximum deadlifts by gender and decade",
       subtitle = "Deadlifting ability hasn't changed much within gender across decades",
       x = "Decade", 
       y = "Maximum deadlift (kgs)") +
  theme_ipsum() +
  theme(legend.position = "top")
  
```

As we would expect, male lifters are able to lift higher maximal weights than women. There isn't any trend in maximum weights lifted over the time-span of this data set in men or women. The distributions by decade clearly show the increasing data available in more recent decads. 

## Conclusions

The `tidybayes` package has a number of ggplot-compatible functions which are perfect for visualising uncertainty and distributions. The development version available on Github remains the version to use for this purpose since most of the most interesting functions remain in development. 

