---
title: Propensity matching
author: Jay Achar
date: '2020-01-26'
slug: propensity-matching
categories:
  - rstats
tags:
  - R
  - analysis
description: 'Analysing observational data'
images:
  - ''
linktitle: ''
type: post
draft: true
---
```{r setup, include = FALSE}
library(tidyverse)

```

## What is propensity matching for? 

Propensity matching is a technique used primarily for the analysis of observational study data. In non-randomised studies, treatment is frequently allocated based on certain characteristics of participants. For example, where a drug is considered to cause more side-effects, it may not be given to older participants by clinicians. This is a concept called **confounding** where a third variable is associated with both the outcome and the explanatory variable of a study question. Participant age may be a confounding characteristic in a study investigating the effect of an HIV treatment regimen on viral suppression for example. 

## Example data

```{r read-data}
# Read in data
mort <- read.csv(file = "https://jay.rbind.io/data/propensity_data.csv")
head(mort)
```



## Additional resources

http://web.hku.hk/~bcowling/examples/propensity.htm